Effectful Semantics for Substructural SSA

### 2023-11-06 pldi24/paper.tex
Tap

@@

\TODO{\(\implies\) E-graph optimization}

\TODO{rewriting substitution + weakening?}

@@

\TODO{this...}

@@

\TODO{Ye Olde Trace Monad}

\TODO{Is Elgot}

\TODO{Ye Olde Nondeterministic Trace Monad}

\TODO{Is Elgot}

\TODO{Ye Olde Reader/Writer/State Transformer}

\TODO{Is Elgot}

\TODO{already done above?}

@@

\TODO{Pomsets...}

@@

\TODO{The SC Monad, which, via composition w/ the state transformer, lets us}

@@

\TODO{Karoubi envelope as premonoidal demonstration?}

@@

>>>>

% Math
\newcommand{\nats}{\mathbb{N}}

@@

@@

\TODO{this:}

\TODO{substitution \(\implies\): loop hoisting, global value numbering, some dead code elimination}

\TODO{rewriting \(\implies\) strength reduction, E-graph optimization, see \cite{cranelift}}

\TODO{inlining}

\TODO{loop fusion}

\TODO{unreachable code elimination}

\TODO{more dead code elimination}

@@

We begin by defining the monad of \textit{nondeterministic traces} generated by a monoid \(M\) and a set \(I\) on \(M\) equipped with an "infinitary action" as follows:
\begin{definition}[Infinitary Action]
  An \textbf{infinitary action} of a monoid \(M\) on a set \(I\) is a pair \((\Sigma, \cdot)\) where \(\Sigma: (\nats \to M) \to I\) is a map from streams of \(M\) to \(I\) and \(\cdot: M \to I \to I\) is a monoid action such that
  \(
    m \cdot \Sigma \ell = \Sigma (m: \ell) 
  \)
\end{definition}
Some examples include:
\begin{itemize}
  \item Any monoid \(M\) equipped with the unique maps \(\Sigma: (\nats \to M) \to \mb{1}\), \(\cdot: M \to \mb{1} \to \mb{1}\)
  \item The monoid of lists of \(A\) equipped with the concatenation map \(\Sigma: (\nats \to A^*) \to A^\omega\) and the concatenation action \(\cdot: A^* \to A^\omega \to A^\omega\)
\end{itemize}
\TODO{better syntax, think how to write \(A^\omega\)...}

\begin{definition}[Trace Monad]
  Given an infinitary action \((\Sigma, \cdot)\) of \(M\) on \(I\), we can define the \textbf{trace monad} \(\ms{Trace}\;\Sigma\;A = A \times M + I\), with
  \TODO{bind}
  \TODO{pure}
\end{definition}

\TODO{is Elgot...}

\begin{definition}[Nondeterministic Trace Monad]
  Given an infinitary action \((\Sigma, \cdot)\) of \(M\) on \(I\), we can define the \textbf{nondeterministic trace monad} \(\ms{Trace}\;\Sigma\;A = \mc{P}^+(A \times M + I)\), with
  \TODO{bind}
  \TODO{pure}
\end{definition}

\TODO{is Elgot...}

\TODO{induces a substructural effectful category...}

@@

\begin{definition}[Pomset] \
  \ \TODO{this}
\end{definition}

@@

\TODO{The SC Monad, which, via composition w/ the state transformer w/ a buffer as state, lets us}

@@

\begin{definition}[Monadic Predicate] \
  \TODO{this and naming}
\end{definition}

\TODO{Proper Weak Memory Monad, preserves the Elgot structure (yey)}

@@

\TODO{Name for ND trace monad}
